---
title: "Eq"
author: "Scott Michael Dickson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Eq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r intro, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
#library(readr)
#library(magrittr)
library(dplyr)
#library(lubridate)
#library(stringr)
#library(grid)
library(ggplot2)
#library(leaflet)
library(Eq)
```

# Eq package overview

The Eq package contains a series of functions that enable the user to quickly load and visualise earthquake data. 

The functions contained within this package are as follows:

* eq_read_data
* eq_clean_data
* eq_location_clean
* geom_timeline
* geom_timeline_label
* eq_map
* eq_create_label

The following code can be used to install the package:

```{r install_ex, eval = FALSE}
library(devtools)
install_github("Eq")
library(Eq)
```

## eq_read_data

The eq_read_data function enables users to read earthquake data from the National Oceanographic and Atmospheric Administration (NOAA) text file.
The input to the function is the path, name, and extension of the csv file containing the data. 
Only the name and extension are required if the file is stored in the current directory.
The output is a data frame. 
The examples below use the data that is included with this package that can be accessed using the following code:

* system.file("extdata", "signif.txt", package = "Eq").
 
Examples:
```{r read_ex}
x <- system.file("extdata", "signif.txt", package = "Eq")
y <- eq_read_data(x)
y
```

## eq_clean_data

The eq_clean_data function enables users to tidy and subset earthquake data.
The columns that are kept in the data are: YEAR, MONTH, DAY, LATITUDE, LONGITUDE, LOCATION_NAME, TOTAL_DEATHS, COUNTRY, EQ_PRIMARY.
Records with negative year values are excluded.
A date column is created from the YEAR, MONTH, and DAY columns.
The input to the function is the name of the input data frame on which to apply the function.
The output is a data frame. 
 
#### Examples:
<!-- print code as text without running it --> 
Code:
```{r clean_ex1, eval = FALSE}
x <- system.file("extdata", "signif.txt", package = "Eq")
x <- eq_read_data(x)
y <- eq_clean_data(x)
y[,c("COUNTRY", "LATITUDE", "LONGITUDE", "DATE", "EQ_PRIMARY", "TOTAL_DEATHS", "LOCATION_NAME", "YEAR", "MONTH", "DAY")]
```

<!-- run code but avoid printing anything -->
```{r clean_ex2, echo = FALSE, results = 'hide'}
x <- system.file("extdata", "signif.txt", package = "Eq")
x <- suppressMessages(eq_read_data(x))
y <- eq_clean_data(x)
y[,c("COUNTRY", "LATITUDE", "LONGITUDE", "DATE", "EQ_PRIMARY", "TOTAL_DEATHS", "LOCATION_NAME", "YEAR", "MONTH", "DAY")]
```

<!-- print results only - avoid reprinting results from previous part of pipe chain -->
Result:
```{r clean_ex3, echo = FALSE}
y[,c("COUNTRY", "LATITUDE", "LONGITUDE", "DATE", "EQ_PRIMARY", "TOTAL_DEATHS", "LOCATION_NAME", "YEAR", "MONTH", "DAY")]
```

## eq_location_clean

The eq_location_clean function enables users to tidy the LOCATION_NAME field in the earthquake data.
It is assumed that country names are followed by a colon when a subsequent city name is provided.
Where only the country name is given in the LOCATION_NAME field, the country name is removed since this information is in the country field.
The LOCATION_NAME field is converted to title case
The input to the function is the name of the input data frame on which to apply the function.
The output is a data frame. 
 
#### Examples:
<!-- print code as text without running it --> 
Code:
```{r clean_loc_ex, eval = FALSE}
# Load either magrittr or dplyr package to allow use of pipe operator %>%
library(dplyr)
x <- system.file("extdata", "signif.txt", package = "Eq")
x <- x %>% 
eq_read_data() %>%
eq_clean_data()
y <- eq_location_clean(x)
y[,c("LOCATION_NAME")]
```

<!-- run code but avoid printing anything -->
```{r clean_loc_ex2, echo = FALSE, results = 'hide'}
x <- system.file("extdata", "signif.txt", package = "Eq")
x <- suppressMessages(eq_read_data(x))
x <- eq_clean_data(x)
y <- eq_location_clean(x)
y[,c("LOCATION_NAME")]
```

<!-- print results only - avoid reprinting results from previous part of pipe chain -->
<!-- print desired columns using syntax df[,c("A","B","E")] or equivalently df[,c(1,2,5)] -->
Result:
```{r clean_loc_ex3, echo = FALSE}
y[,c("LOCATION_NAME")]
```

## geom_timeline

The geom_timeline function enables users to plot a timeline of the earthquake data.
Earthquakes between dates xmin and xmax are plotted.
Separate timelines can be shown for different groups within a single plot e.g. using the COUNTRY field.
The size of the points on the timeline can be set to distinguish earthquakes of differing magnitude e.g using the EQ_PRIMARY field.
The fill colour of the points on the timeline can be set to distinguish earthquakes of differing number of deaths e.g using the TOTAL_DEATHS field.  
The output is a timeline showing eathquake sizes and intensities between specified dates. 
 
#### Examples:
<!-- print code as text without running it --> 
Code:
```{r timeline_ex1a, eval = FALSE}
# Single timeline of all earthquakes in date range from all countries

# Load either magrittr or dplyr package to allow use of pipe operator %>%
# Load ggplot2 package to allow use of ggplot function
library(dplyr)
library(ggplot2)
x1 <- system.file("extdata", "signif.txt", package = "Eq")
x1 <- x1 %>%
eq_read_data() %>%
eq_clean_data() %>%
eq_location_clean()
y1 <- ggplot(data = x1) +
geom_timeline(aes(	x = DATE,
					xmin = as.Date("2013-01-01"),
					xmax = as.Date("2015-12-31"),
					size = EQ_PRIMARY,
					fill = TOTAL_DEATHS          		
)) +
labs(size = "Richter scale value", fill = "# deaths") +
theme(legend.position="bottom")
y1
```

<!-- run code but avoid printing anything -->
```{r timeline_ex1b, echo = FALSE, results = 'hide'}
x1 <- system.file("extdata", "signif.txt", package = "Eq")
x1 <- suppressMessages(eq_read_data(x1))
x1 <- eq_clean_data(x1)
x1 <- eq_location_clean(x1)
y1 <- ggplot(data = x1) +
geom_timeline(aes(	x = DATE,
					xmin = as.Date("2013-01-01"),
					xmax = as.Date("2015-12-31"),
					size = EQ_PRIMARY,
					fill = TOTAL_DEATHS          		
)) +
labs(size = "Richter scale value", fill = "# deaths") +
theme(legend.position="bottom")
```

<!-- print results only - avoid reprinting results from previous part of pipe chain -->
<!-- use fig.width and fig.height to adjust figure sizes (inches) e.g. fig.width = 4 -->
Result:
```{r timeline_ex1c, echo = FALSE, fig.width = 7, fig.height = 3}
y1
```

<!-- print code as text without running it --> 
Code:
```{r timeline_ex2a, eval = FALSE}
# Separate timelines for earthquakes in each country

# Load either magrittr or dplyr package to allow use of pipe operator %>%
# Load ggplot2 package to allow use of ggplot function
library(dplyr)
library(ggplot2)
x2 <- system.file("extdata", "signif.txt", package = "Eq")
x2 <- x2 %>%
eq_read_data() %>%
eq_clean_data() %>%
eq_location_clean()
y2 <- ggplot(data = x2) +
geom_timeline(aes(	x = DATE,
					xmin = as.Date("2000-01-01"),
					xmax = as.Date("2015-12-31"),
					y = COUNTRY,
					size = EQ_PRIMARY,
					fill = TOTAL_DEATHS          		
)) +
labs(size = "Richter scale value", fill = "# deaths") +
theme(legend.position="right")
y2
```

<!-- run code but avoid printing anything -->
```{r timeline_ex2b, echo = FALSE, results = 'hide'}
x2 <- system.file("extdata", "signif.txt", package = "Eq")
x2 <- suppressMessages(eq_read_data(x2))
x2 <- eq_clean_data(x2)
x2 <- eq_location_clean(x2)
y2 <- ggplot(data = x2) +
geom_timeline(aes(	x = DATE,
					xmin = as.Date("2000-01-01"),
					xmax = as.Date("2015-12-31"),
					y = COUNTRY,
					size = EQ_PRIMARY,
					fill = TOTAL_DEATHS          		
)) +
labs(size = "Richter scale value", fill = "# deaths") +
theme(legend.position="right")
```

<!-- print results only - avoid reprinting results from previous part of pipe chain -->
Result:
```{r timeline_ex2c, echo = FALSE, fig.width = 7, fig.height = 40}
y2
```

<!-- print code as text without running it --> 
Code:
```{r timeline_ex3a, eval = FALSE}
# Separate timelines for earthquakes in selected countries

# Load either magrittr or dplyr package to allow use of pipe operator %>%
# Load ggplot2 package to allow use of ggplot function
library(dplyr)
library(ggplot2)
x3 <- system.file("extdata", "signif.txt", package = "Eq")
x3 <- x3 %>%
eq_read_data() %>%
eq_clean_data() %>%
eq_location_clean()
y3 <- ggplot(data = subset(x3, COUNTRY %in% c("USA" , "CHINA"))) +
geom_timeline(aes(	x = DATE,
					xmin = as.Date("2000-01-01"),
					xmax = as.Date("2015-12-31"),
					y = COUNTRY,
					size = EQ_PRIMARY,
					fill = TOTAL_DEATHS          		
)) +
labs(size = "Richter scale value", fill = "# deaths") +
theme(legend.position="right")
y3
```

<!-- run code but avoid printing anything -->
```{r timeline_ex3b, echo = FALSE, results = 'hide'}
x3 <- system.file("extdata", "signif.txt", package = "Eq")
x3 <- suppressMessages(eq_read_data(x3))
x3 <- eq_clean_data(x3)
x3 <- eq_location_clean(x3)
y3 <- ggplot(data = subset(x3, COUNTRY %in% c("USA" , "CHINA"))) +
geom_timeline(aes(	x = DATE,
					xmin = as.Date("2000-01-01"),
					xmax = as.Date("2015-12-31"),
					y = COUNTRY,
					size = EQ_PRIMARY,
					fill = TOTAL_DEATHS          		
)) +
labs(size = "Richter scale value", fill = "# deaths") +
theme(legend.position="right")
```

<!-- print results only - avoid reprinting results from previous part of pipe chain -->
Result:
```{r timeline_ex3c, echo = FALSE, fig.width = 7, fig.height = 4}
y3
```

## geom_timeline_label

The geom_timeline_label function enables users to add location labels to timeline plots of the earthquake data.
Labels can be added to a subset of the plotted earthquakes using the n_max parameter to show labels of the highest magnitude earthquakes.
Parameters other than n_max should be the same values as used in the geom_timeline function.  
The output is a timeline showing eathquake sizes and intensities between specified dates with location name labels. 
 
#### Examples:
<!-- print code as text without running it --> 
Code:
```{r timeline_lbl_ex1a, eval = FALSE}
# Separate timelines for earthquakes in selected countries with labels for highest magnitude earthquakes only

# Load either magrittr or dplyr package to allow use of pipe operator %>%
# Load ggplot2 package to allow use of ggplot function
library(dplyr)
library(ggplot2)
x1 <- system.file("extdata", "signif.txt", package = "Eq")
x1 <- x1 %>%
eq_read_data() %>%
eq_clean_data() %>%
eq_location_clean()
y1 <- ggplot(data = x1) +
geom_timeline(aes(	x = DATE,
					xmin = as.Date("2013-01-01"),
					xmax = as.Date("2015-12-31"),
					size = EQ_PRIMARY,
					fill = TOTAL_DEATHS          		
)) + 
geom_timeline_label(aes(x = DATE,
						xmin = as.Date("2013-01-01"),
						xmax = as.Date("2015-12-31"), 
						label = LOCATION_NAME, 
						size = EQ_PRIMARY), 
					n_max = 5) +
labs(size = "Richter scale value", fill = "# deaths") +
theme(legend.position="bottom")
y1
```

<!-- run code but avoid printing anything -->
```{r timeline_lbl_ex1b, echo = FALSE, results = 'hide'}
x1 <- system.file("extdata", "signif.txt", package = "Eq")
x1 <- suppressMessages(eq_read_data(x1))
x1 <- eq_clean_data(x1)
x1 <- eq_location_clean(x1)
y1 <- ggplot(data = x1) +
geom_timeline(aes(	x = DATE,
					xmin = as.Date("2013-01-01"),
					xmax = as.Date("2015-12-31"),
					size = EQ_PRIMARY,
					fill = TOTAL_DEATHS          		
)) + 
geom_timeline_label(aes(x = DATE,
						xmin = as.Date("2013-01-01"),
						xmax = as.Date("2015-12-31"), 
						label = LOCATION_NAME, 
						size = EQ_PRIMARY), 
					n_max = 5) +
labs(size = "Richter scale value", fill = "# deaths") +
theme(legend.position="bottom")
```

<!-- print results only - avoid reprinting results from previous part of pipe chain -->
Result:
```{r timeline_lbl_ex1c, echo = FALSE, fig.width = 7, fig.height = 4}
y1
```

<!-- print code as text without running it --> 
Code:
```{r timeline_lbl_ex3a, eval = FALSE}
# Separate timelines for earthquakes in selected countries with labels for highest magnitude earthquakes only

# Load either magrittr or dplyr package to allow use of pipe operator %>%
# Load ggplot2 package to allow use of ggplot function
library(dplyr)
library(ggplot2)
x3 <- system.file("extdata", "signif.txt", package = "Eq")
x3 <- x3 %>%
eq_read_data() %>%
eq_clean_data() %>%
eq_location_clean()
y3 <- ggplot(data = subset(x3, COUNTRY %in% c("USA" , "CHINA"))) +
geom_timeline(aes(	x = DATE,
					xmin = as.Date("2011-01-01"),
					xmax = as.Date("2016-12-31"),
					y = COUNTRY,
					size = EQ_PRIMARY,
					fill = TOTAL_DEATHS          		
)) + 
geom_timeline_label(aes(x = DATE,
						xmin = as.Date("2011-01-01"),
						xmax = as.Date("2016-12-31"),
						y = COUNTRY, 
						label = LOCATION_NAME, 
						size = EQ_PRIMARY), 
					n_max = 4) +
labs(size = "Richter scale value", fill = "# deaths") +
theme(legend.position="bottom")
y3
```

<!-- run code but avoid printing anything -->
```{r timeline_lbl_ex3b, echo = FALSE, results = 'hide'}
x3 <- system.file("extdata", "signif.txt", package = "Eq")
x3 <- suppressMessages(eq_read_data(x3))
x3 <- eq_clean_data(x3)
x3 <- eq_location_clean(x3)
y3 <- ggplot(data = subset(x3, COUNTRY %in% c("USA" , "CHINA"))) +
geom_timeline(aes(	x = DATE,
					xmin = as.Date("2011-01-01"),
					xmax = as.Date("2016-12-31"),
					y = COUNTRY,
					size = EQ_PRIMARY,
					fill = TOTAL_DEATHS          		
)) + 
geom_timeline_label(aes(x = DATE,
						xmin = as.Date("2011-01-01"),
						xmax = as.Date("2016-12-31"),
						y = COUNTRY, 
						label = LOCATION_NAME, 
						size = EQ_PRIMARY), 
					n_max = 4) +
labs(size = "Richter scale value", fill = "# deaths") +
theme(legend.position="bottom")
```

<!-- print results only - avoid reprinting results from previous part of pipe chain -->
Result:
```{r timeline_lbl_ex3c, echo = FALSE, fig.width = 7, fig.height = 9}
y3
```

## eq_map

The eq_map function creates an interactive map of earthquake data.
An input to the function is the name of the input data frame on which to apply the function.
The annot_col parameter can be used to specify the field to use as a label for points on the map e.g. DATE.
The output is an interactive map.
 
#### Examples:
<!-- print code as text without running it --> 
Code:
```{r map_ex1, eval = FALSE}
# Load either magrittr or dplyr package to allow use of pipe operator %>%
library(dplyr)
x <- system.file("extdata", "signif.txt", package = "Eq")
x <- x %>%
eq_read_data()
eq_clean_data()
eq_location_clean()
y <- x %>% 
dplyr::filter(COUNTRY == "MEXICO" & lubridate::year(DATE) >= 2000) %>% 
eq_map(annot_col = "DATE")
y
```

<!-- run code but avoid printing anything -->
```{r map_ex2, echo = FALSE, results = 'hide'}
x <- system.file("extdata", "signif.txt", package = "Eq")
x <- suppressMessages(eq_read_data(x))
x <- eq_clean_data(x)
x <- eq_location_clean(x)
y <- x %>% 
dplyr::filter(COUNTRY == "MEXICO" & lubridate::year(DATE) >= 2000) %>% 
eq_map(annot_col = "DATE")
y
```

<!-- print results only - avoid reprinting results from previous part of pipe chain -->
Result:
```{r map_ex3, echo = FALSE, fig.width = 7, fig.height = 4}
y
```

## eq_create_label

The eq_create_label function creates creates and applies formatting to the labels for the points on the interactive map of earthquake data.
The input to the function is the name of the input data frame on which to apply the function.
 
#### Examples:
<!-- print code as text without running it --> 
Code:
```{r map_lbl_ex1, eval = FALSE}
x <- system.file("extdata", "signif.txt", package = "Eq")
x <- x %>% 
eq_read_data()
eq_clean_data()
eq_location_clean()
y <- x %>% 
dplyr::filter(COUNTRY == "MEXICO" & lubridate::year(DATE) >= 2000) %>%
dplyr::mutate(popup_text = eq_create_label(.)) %>% 
eq_map(annot_col = "popup_text")
y
```

<!-- run code but avoid printing anything -->
```{r map_lbl_ex2, echo = FALSE, results = 'hide'}
x <- system.file("extdata", "signif.txt", package = "Eq")
x <- suppressMessages(eq_read_data(x))
x <- eq_clean_data(x)
x <- eq_location_clean(x)
y <- x %>% 
dplyr::filter(COUNTRY == "MEXICO" & lubridate::year(DATE) >= 2000) %>%
dplyr::mutate(popup_text = eq_create_label(.)) %>% 
eq_map(annot_col = "popup_text")
y
```

<!-- print results only - avoid reprinting results from previous part of pipe chain -->
Result:
```{r map_lbl_ex3, echo = FALSE, fig.width = 7, fig.height = 4}
y
```

